---
title: "DSC 532- Fuel Regression project"
output: pdf_document
date: "2025-04-01"
---

Dataset Overview
Observations: 1,067 rows (vehicles)
Variables: 13 columns
Purpose: Contains fuel consumption and CO2 emissions data for vehicles from the year 2014




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(moments)
library(nortest)
library(corrplot)
library(caret)
library(tidyr)

```
Load the FuelConsumption dataset and preview the first 6 rows of the dataset.

```{r load-data, message=FALSE}
data <- read.csv("FuelConsumption.csv")
head(data)
```

The dataset includes features like:

```{r}
colnames(data)
``` 
Data types and Structure of the dataset.
```{r}
str(data)
```


Statistical summary of numerical variables.

From the summary table we can observe that vehicles with larger engines and more cylinders tend to have higher fuel consumption and CO₂ emissions.

Also,there's a wide variety of fuel efficiency, from very eco-friendly models to high-emission vehicles.

```{r}
summary(data)
```
Statistical summary of numerical variables.

From the summary table we can observe that vehicles with larger engines and more cylinders tend to have higher fuel consumption and CO₂ emissions.

Also,there's a wide variety of fuel efficiency, from very eco-friendly models to high-emission vehicles.

From the model year column we can observe that all the cars are from 2014

 
```{r}

colSums(is.na(data))  # This gives you count of NAs per column

```

We have 0 missing values, so we don’t need to impute or remove anything.

```{r}

# Number of duplicated rows
sum(duplicated(data))

```

There are no duplicated rows in the dataset

```{r}

# Check unique values for categorical columns
cat_cols <- names(data)[sapply(data, is.character)]

# Apply unique() to each categorical column and name the output
unique_values <- lapply(data[cat_cols], unique)

# Display the unique values nicely
unique_values


```

The dataset contains a rich variety of categorical features, including 39 car makes, hundreds of models, 16 vehicle classes, multiple transmission types, and 4 fuel types.

All categorical columns are free of spelling errors, case inconsistencies, or data entry mistakes.


## Including Plots

You can also embed plots, for example:
library(tidyr)
```{r wide-histogram, fig.width=10, fig.height=6}
numeric_vars <- data %>% select(ENGINESIZE, CYLINDERS, FUELCONSUMPTION_COMB, CO2EMISSIONS)

numeric_vars %>%
  gather() %>%
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free", ncol = 2) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Key Numeric Variables")
```

```{r}
data %>%
  count(MAKE, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(MAKE, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Car Makes", x = "Make", y = "Count")

```

```{r}

data %>%
  count(MODEL, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(MODEL, n), y = n)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Car Models", x = "Model", y = "Count")
```
```{r}
ggplot(data, aes(x = VEHICLECLASS, fill = VEHICLECLASS)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Vehicle Classes", x = "Vehicle Class", y = "Count") +
  theme(legend.position = "none")
```




### 📊 Categorical Variable Distributions

```{r categorical-plots, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}

# 1. Top 10 Car Makes
library(dplyr)
library(ggplot2)

data %>%
  count(MAKE, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(MAKE, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Car Makes", x = "Make", y = "Count")

# 2. Top 10 Car Models
data %>%
  count(MODEL, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(MODEL, n), y = n)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Car Models", x = "Model", y = "Count")

# 3. Vehicle Class Distribution
ggplot(data, aes(x = VEHICLECLASS, fill = VEHICLECLASS)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Vehicle Classes", x = "Vehicle Class", y = "Count") +
  theme(legend.position = "none")

# 4. Fuel Type Distribution
ggplot(data, aes(x = FUELTYPE, fill = FUELTYPE)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Fuel Type Distribution", x = "Fuel Type", y = "Count") +
  theme(legend.position = "none")

# 5. Transmission Code Distribution
ggplot(data, aes(x = TRANSMISSION, fill = TRANSMISSION)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Transmission Code Distribution", x = "Transmission", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")
```









CO2EMISSIONS: Most values are concentrated between 150 and 300 g/km, with a gradual decline toward higher emissions. This suggests that have a slightly skewed distribution.

CYLINDERS: The distribution is clearly categorical, with strong peaks at 4, 6, and 8 cylinders—indicating these are the dominant engine types in the dataset.

ENGINESIZE: Smaller engines (2 to 3.5 liters) are more common, while larger engines appear less frequently. The data is not perfectly symmetrical, showing some right skew.

FUELCONSUMPTION_COMB: The majority of vehicles consume around 8 to 12 liters per 100 km. Fewer vehicles fall into the higher consumption range, again showing a right-skewed shape.

```{r pressure, echo=FALSE}
# 1. MPG Difference Check Without Creating a Column
calculated_mpg <- 235.214583 / data$FUELCONSUMPTION_COMB
mpg_diff <- data$FUELCONSUMPTION_COMB_MPG - calculated_mpg

# Threshold for the top 5% largest differences
percentile_95 <- quantile(mpg_diff, 0.95)
print(percentile_95)

# Count how many exceed it
sum(mpg_diff > percentile_95)  # Still gives 52


# 2. Fuel Consumption Check (Simple and Weighted Average)

# Simple average
simple_avg <- (data$FUELCONSUMPTION_CITY + data$FUELCONSUMPTION_HWY) / 2
# Weighted average
weighted_avg <- 0.55 * data$FUELCONSUMPTION_CITY + 0.45 * data$FUELCONSUMPTION_HWY

# Compare to actual
diff_simple <- abs(data$FUELCONSUMPTION_COMB - simple_avg)
diff_weighted <- abs(data$FUELCONSUMPTION_COMB - weighted_avg)

# Mean differences
mean(diff_simple)
mean(diff_weighted)

# Use IQR method for threshold
Q1 <- quantile(diff_weighted, 0.25)
Q3 <- quantile(diff_weighted, 0.75)
IQR <- Q3 - Q1
threshold <- Q3 + 1.5 * IQR
print(threshold)

# Check how many values are above threshold
sum(diff_weighted > threshold)

# Or use 0.1 as your cutoff
sum(diff_weighted > 0.1)
```



```{r}
library(stringr)



# Start with NA
data$TransmissionGroup <- NA

# Add all transmission groupings
data$TransmissionGroup[grepl("^A[0-9]$", data$TRANSMISSION)] <- "Automatic"
data$TransmissionGroup[grepl("^AM[0-9]$", data$TRANSMISSION)] <- "Automated Manual"
data$TransmissionGroup[grepl("^AS[0-9]$", data$TRANSMISSION)] <- "Select Shift Auto"
data$TransmissionGroup[grepl("^AV", data$TRANSMISSION)] <- "CVT"
data$TransmissionGroup[grepl("^M[0-9]$", data$TRANSMISSION)] <- "Manual"  # NEW

# Convert to factor
data$TransmissionGroup <- factor(dataset$TransmissionGroup)

# Check if any values still NA
table(data$TransmissionGroup, useNA = "ifany")

head(data)

sum(is.na(data$TransmissionGroup))
```

```{r pressure, echo=FALSE}
ggplot(data %>% filter(!is.na(TransmissionGroup)), aes(x = TransmissionGroup, y = CO2EMISSIONS, fill = TransmissionGroup)) +
  geom_boxplot() +
  theme_minimal()

```

```{r}
table(data$TransmissionGroup, useNA = "ifany")
sum(is.na(data$TransmissionGroup))
```








```{r scatter-enginesize}
# Scatterplots with trend lines
ggplot(data, aes(x = ENGINESIZE, y = CO2EMISSIONS)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "CO2 Emissions vs Engine Size", x = "Engine Size (L)", y = "CO2 Emissions (g/km)")
```

There is a clear positive relationship. As fuel consumption increases, CO2 emissions also increase. The points are close to the line, showing a strong connection.





```{r pressure, echo=FALSE}
ggplot(data, aes(x = FUELCONSUMPTION_COMB, y = CO2EMISSIONS)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "darkgreen") +
  theme_minimal() +
  labs(title = "CO2 Emissions vs Combined Fuel Consumption")
```

This plot also shows a positive trend. Vehicles with bigger engines usually produce more CO2, but the points are more spread out, so the relationship is not as strong as with fuel consumption.




```{r pressure, echo=FALSE}
ggplot(data, aes(x = FUELTYPE, y = CO2EMISSIONS, fill = FUELTYPE)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "CO2 Emissions by Fuel Type", x = "Fuel Type", y = "CO2 Emissions (g/km)") +
  theme(legend.position = "none")  

```

CO2 Emissions by Fuel Type:

Vehicles using fuel type E tend to have the highest CO2 emissions based on the median.
Fuel type D shows the lowest median emissions.
All fuel types show some variation, but type E also has a wider range and more high outliers.

```{r pressure, echo=FALSE}
install.packages("GGally")
library(GGally)
library(dplyr)

ggpairs(data %>% select(ENGINESIZE, CYLINDERS, FUELCONSUMPTION_COMB, CO2EMISSIONS))

```



```{r pressure, echo=FALSE}
library(corrplot)
library(dplyr)
# Only numeric columns
cor_data <- data %>% select_if(is.numeric)
cor_matrix <- cor(cor_data)

# Plot
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8, title = "Correlation Matrix", mar=c(0,0,1,0))
```



Dark blue = strong positive correlation (close to +1)
Dark red = strong negative correlation (close to -1)
Lighter colors = weak or no correlation (close to 0)

CO2EMISSIONS is strongly linked to:
FUELCONSUMPTION_COMB
FUELCONSUMPTION_CITY
FUELCONSUMPTION_HWY
ENGINESIZE and CYLINDERS

FUELCONSUMPTION_COMB_MPG is negatively related to CO2 (more MPG = less emissions).

MODELYEAR has no clear connection with the other variables.


Simple Linear Regression

```{r pressure, echo=FALSE}
# Fit the model
model_simple <- lm(CO2EMISSIONS ~ FUELCONSUMPTION_COMB, data = data)

# See the summary
summary(model_simple)
```

```{r pressure, echo=FALSE}
library(ggplot2)
ggplot(data, aes(x = FUELCONSUMPTION_COMB, y = CO2EMISSIONS)) +
  geom_point(color = "black", alpha = 0.6) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Simple Linear Regression: CO2 vs Fuel Consumption",
       x = "Combined Fuel Consumption (L/100km)",
       y = "CO2 Emissions (g/km)") +
  theme_minimal()
```
```{r pressure, echo=FALSE}
# Simple linear regression: CO2 vs Engine Size
model_engine <- lm(CO2EMISSIONS ~ ENGINESIZE, data = data)

# Show the summary
summary(model_engine)
```

```{r pressure, echo=FALSE}
ggplot(data, aes(x = ENGINESIZE, y = CO2EMISSIONS)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Simple Linear Regression: CO2 vs Engine Size",
       x = "Engine Size (L)",
       y = "CO2 Emissions (g/km)") +
  theme_minimal()
```



Both variables are strong predictors of CO2 emissions.
The model with FUELCONSUMPTION_COMB explains slightly more of the variation (R² = 0.796 vs 0.764).
Residuals are smaller in the first model, which also suggests a better fit.
Both models are statistically significant and show clear positive trends.
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.




#Model Evaluation 
```{r pressure, echo=FALSE}
library(ggplot2)
model_multi <- lm(CO2EMISSIONS ~ ENGINESIZE + CYLINDERS + FUELCONSUMPTION_COMB , data = data)
summary(model_multi)

library(caret)

set.seed(123)
split <- createDataPartition(data$CO2EMISSIONS, p = 0.8, list = FALSE)
train <- data[split, ]
test <- data[-split, ]

model_final <- lm(CO2EMISSIONS ~ ENGINESIZE + CYLINDERS + FUELCONSUMPTION_COMB , data = train)
predictions <- predict(model_final, test)

# RMSE
rmse <- sqrt(mean((predictions - test$CO2EMISSIONS)^2))
rmse

#We evaluate the model by splitting the data into training and test sets (80/20). We then compare actual vs predicted values and inspect residuals.
``` 

```{r pressure, echo=FALSE}
# Plot: Actual vs Predicted
ggplot(data.frame(Actual = test$CO2EMISSIONS, Predicted = predictions),
       aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  theme_minimal() +
  labs(title = "Actual vs Predicted CO2 Emissions",
       x = "Actual CO2",
       y = "Predicted CO2")
```




```{r pressure, echo=FALSE}
residuals <- test$CO2EMISSIONS - predictions

ggplot(data.frame(Residuals = residuals), aes(x = Residuals)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Residuals",
       x = "Residuals (Actual - Predicted)",
       y = "Count")


```


```{r pressure, echo=FALSE}
ggplot(data.frame(Residuals = residuals), aes(x = "", y = Residuals)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(title = "Boxplot of Residuals",
       x = "",
       y = "Residuals (Actual - Predicted)")

```





Conclusion

Which variables were most important?
The most important variable in predicting CO2 emissions was FUELCONSUMPTION_COMB, which had the highest correlation and produced the strongest simple regression model. Other useful predictors included ENGINESIZE, CYLINDERS, and TransmissionType.

Was the model good?
Yes, the final multiple regression model performed well. It explained a large part of the variation in CO2 emissions (high R² value), and its predictions were close to the actual values in the test set. The residuals were centered around zero with a few expected outliers, showing the model was reasonably accurate.
